### 2016-03-01 paper/icfp16-submission/semantics.tex
proofs

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
is $\typeof{\subst{\su}{\ehole{}}} \rightarrow \typeof{e}$.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each successful evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at the last step.
  %
  An examination of the rules shows that only place this happens is
  in the first case of \forcesym.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} \rightarrow \_ = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
% \begin{description}
  %\item
  [Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  %\item
  [Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
%\end{description}
\end{proof}

>>>>

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input.
%
We can extract this type from evaluation traces.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\typeof{\subst{\su}{\ehole{}}}$.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a successful call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at the last step.
  %
  An examination of the rules shows that only place this happens is
  in the second case of \forcesym.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  \end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
\end{description}
\end{proof}
